master thesis

### 2018-03-21 Paper - Latex/Paper - Skizze - v4.tex
Paper - Results - BIC of unknown states

        First traverse from the leaf nodes to the root / move down the tree / postorder tree traversal.
          \begin{wrapfigure}{r}{0.5\textwidth}
            \centering
            \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
            \caption{Fitch algorithm for binary trees. \\
              The unknown leaf node is discribed with both states. Computed internal nodes (exclusive the 
              root node) consists of three sets, where the last set is the final one (bold). \\
              From the second internal node (seen from the root node) there are several possibilities to 
              create the second and third set.}
            \label{fig: binary Fitch}
          \end{wrapfigure} 
          Each internal node is formed from its child nodes, where at the beginning the only information
          lies. \\
        Second traverse from the root node to the leafs. Each internal node is formed from its father node 
          and its sibling node. \\
        Last traversion (direction does not matter): Build the final state for every node. It is formed 
          from the sets of previous traversals. \\
        (The original Fitch algorithm was designed to minimize transitions without predicting actual states 
          of internal nodes, so it was just the first traversal.) \\
        The extension to the non-binary case is quite obvious, but holds some opportunities. In this case, 
          more than two children may be present for the first traversal, but the incision or union may 
          also be formed over more than two sets. Also in the second traversing, there may be several 
          sibling nodes. However, there are several possibilities here that were all tested and compared 
          in the simulation. Some of these options are already available in the binary case:
        \begin{itemize}
          \item The father node has (except for the root node) two state sets, because he came through 
            the up-traversing previously. Are both sets used or only the first traversing?
          \item Since there are several siblings, do \colorbox{red}{you} first of all make the cut or union, 
            or directly in the whole with the father node?
        \end{itemize}
        The first point already has an effect on the binary case. Figure \ref{fig: binary Fitch} shows 
          both possibilities of the three sets. \\
        Cunningham uses only the first state set of the father node \cite{Cunningham1998}. \\
        From these two points four different versions of Fitch were formed:
        \begin{enumerate}
          \item Fitch 1: First state set of father node; intersection/union of siblings first.
          \item Fitch 2: First state set of father node; intersection/union of siblings together with father node.
          \item Fitch 3: Both state sets of father node; intersection/union of siblings first.
          \item Fitch 4: Both state sets of father node; intersection/union of siblings together with father node sets.
        \end{enumerate}
        \begin{figure}
          \centering
          \includegraphics[width=0.8\textwidth]{Figures/simulation_fitch_evaluation.png}
          \caption{Test of Fitch Versions.}
          \label{fig:Fitch versions}
        \end{figure}
        These four versions were tested in the simulation with 100 trees and 10000 leaf nodes and a
          distribution of 60~\% FL to 40~\% P. Figure \ref{fig:Fitch versions} shows this over all unknown 
          node percentage. \\
        At 90~\% unknown nodes and 90~\% of multifurcation of the internal nodes, version 1 was 89.26~\%, 
          version 2 was 89.26~\%, version 3 was 89.35~\%, and version 4 was 89.31~\% correct. Therefore, 
          only version 3 was used for all further simulations.

@@

    We examined the ridge of multifurcation of the tree. A complete phylogenetic tree would be 
      binary, which means the number of leaf nodes is closely to the number of internal nodes. But 
      since we only work with a synthesis tree, this tree is multifurcated: we have 241 974 internal 
      nodes and 2 293 463 leaf nodes. \\
    For a first overview we collected for every node its number of children (degree $-1$), and plotted
      this in two histograms, see figure \ref{fig:childrenOfNodes}. \\
    The multifurcation affects only the internal nodes. We collected the number of children $-2$ of 
      every node (a node with two children is binary). That means it discribes the number of nodes which we have lost from the real (binary) 
    phylogenetic tree. \\
    \begin{figure}[h!]
      \centering
      \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.9\textwidth]{Figures/multifurc.pdf}
        \caption{Histogram with automatic binwidth. \\ ~ \\ ~}
      \end{subfigure}
      ~~~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
        %(or a blank line to force the subfigure onto a new line)
      \begin{subfigure}[b]{0.42\textwidth}
        \includegraphics[trim = 0mm 0mm 30mm 0mm, clip, width=0.9\textwidth]{Figures/multifurc_small.pdf}
        \caption{Histogram with $binwidth = 1$. \\ light blue: binary; \\ dark blue: multifunction}
      \end{subfigure}
      \caption{Histograms about the multifurcation of the internal nodes of the synthesis tree.}
      \label{fig:childrenOfNodes}
    \end{figure}
    As \colorbox{red}{you} can see, we are very far from a binary tree. \\
    Some subtrees have been examined, these have the following percentages of missing information: See 
      table \ref{table:percentage loss information subtrees}.
    \begin{table}[h!]
      \begin{center}
        \begin{tabular}{ |l|r|r||r|r|r| }
          \hline
          Subtree of & unknown states & multifurcation \\ 
          \hline \hline
          Eukaryota       & 97.34~\%  & 98.16~\% \\
          \hline \hline
          Metazoa         & 96.44~\%  & 87.93~\% \\ \hline
          Fungi           & 98.87~\%  & 96.97~\% \\ \hline
          Chloroplastida  & 99.14~\%  & 89.46~\% \\
          \hline \hline            
          % Arthropoda      & 97.49~\%  & 89.95~\% \\ \hline
          Apicomplexa     & 86.26~\%  & 87.16~\% \\ \hline
          Nematoda        & 89.01~\%  & 88.59~\% \\ \hline
          Chordata        & 88.59~\%  & \cellcolor{green!50}66.49~\% \\ \hline
          Platyhelminthes & \cellcolor{green!50}68.73~\%  & 80.34~\% \\
          \hline \hline            
          Insecta         & 97.11~\%  & 90.78~\% \\
          \hline  
        \end{tabular}
      \end{center}
      \caption{Examination of subtrees regarding missing information. \\ 
        % Taxa of subtrees: Domain, Kingdom, Phylum, Class \\ 
        Unknown states: missing state information of leaf nodes \\ 
        Multifurcation: missing internal nodes}
      \label{table:percentage loss information subtrees} 
    \end{table}

@@

      % * Residuals: Fehler - wieviele Werte sind nicht gut modelliert. (umso kleiner umso besser - grün) \\
      The Residuals give us the error of the model. If the value is small, our data will be well modeled. \\
      Within the complexity classes it can be seen that the mean height gives the best additional factor. \\

@@

      Next to the problem of the multifurcation of the tree is the less of data we have for the species.
        For the ancestral state reconstruction, we need information in the leaf nodes. \\
      The eukaryotic synthesis tree has 293 463 leaf nodes. The GloBI database has 5 346 414 interactions 
        (at this timepoint). Out of this data we got 51 337 distinct free-living species and 47 332 
        distinct parasite species $\rightarrow$ unknown nodes 2194794 ($\approx 95.7~\%$). \\
      We found also $57,352$ (not distinct) source species and $809,993$ (not distinct) target
        species without OTT ids. Since we currently use only OTT ids, we could not use this information. \\
      \todo{With this only $\approx 4.3~\%$ information in our leaf nodes are ...} \\
      We also compared different models in terms of their BICs (Table: \ref{table:BIC unknown information}). 
        The Residuals are not very meaningful here, since all models have different dimensions.

      \begin{table}[h!]
        \begin{center}
          \begin{tabular}{ |l|r|r|r| }
            \hline
            Model / Taxa & Kingdom & Phylum & Class \\
            \hline \hline
            multifurc $\sim$ taxa & 545740 & \cellcolor{green!10}499227 & \cellcolor{green!30}482265 \\
            \hline
            multifurc $\sim$ taxa + depth & 544789 & \cellcolor{green!10}493017 & \cellcolor{green!50}478998 \\
            \hline
            multifurc $\sim$ taxa * depth & 544062 & \cellcolor{green!30}488366 & \cellcolor{green!50}476382 \\
            \hline
          \end{tabular}
        \end{center}
        \caption{Residuals of unknown information}
        \label{table:...}
      \end{table}

      \begin{table}[h!]
        \begin{center}
          \begin{tabular}{ |l|r|r|r|r| }
            \hline
            Model / Taxa & Kingdom & Phylum & Class & Order \\
            \hline \hline
            multifurc $\sim$ taxa & \cellcolor{green!15}545799 & \cellcolor{green!35}500004 & \cellcolor{green!45}485121 & XXXXX \\
            \hline
            multifurc $\sim$ taxa + depth & \cellcolor{green!15}544862 & \cellcolor{green!40}493808 & \cellcolor{green!45}481869 & \cellcolor{green!50}478851 \\
            \hline
            multifurc $\sim$ taxa * depth & \cellcolor{green!15}544179 & \cellcolor{green!45}489845 & \cellcolor{green!45}481494 & \cellcolor{green!50}478188 \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{BICs of unknown information}
        \label{table:BIC unknown information} 
      \end{table}

      \todo{Taxa like order or family were too expensive to calculate...}

@@

>>>>

        First traverse from the leaf nodes to the root / move down the tree / postorder tree traversal.
          \begin{wrapfigure}{r}{0.5\textwidth}
            \centering
            \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
            \caption{Fitch algorithm for binary trees. \\
              The unknown leaf node is discribed with both states. Computed internal nodes (exclusive the 
              root node) consists of three sets, where the last set is the final one (bold). \\
              From the second internal node (seen from the root node) there are several possibilities to 
              create the second and third set.}
            \label{fig: binary Fitch}
          \end{wrapfigure} 
          Each internal node is formed from its child nodes, where at the beginning the only information
          lies. \\
        Second traverse from the root node to the leafs. Each internal node is formed from its father node 
          and its sibling node. \\
        Last traversion (direction does not matter): Build the final state for every node. It is formed 
          from the sets of previous traversals. \\
        (The original Fitch algorithm was designed to minimize transitions without predicting actual states 
          of internal nodes, so it was just the first traversal.) \\
        The extension to the non-binary case is quite obvious, but holds some opportunities. In this case, 
          more than two children may be present for the first traversal, but the incision or union may 
          also be formed over more than two sets. Also in the second traversing, there may be several 
          sibling nodes. However, there are several possibilities here that were all tested and compared 
          in the simulation. Some of these options are already available in the binary case:
        \begin{itemize}
          \item The father node has (except for the root node) two state sets, because he came through 
            the up-traversing previously. Are both sets used or only the first traversing?
          \item Since there are several siblings, do \colorbox{red}{you} first of all make the cut or union, 
            or directly in the whole with the father node?
        \end{itemize}
        The first point already has an effect on the binary case. Figure \ref{fig: binary Fitch} shows 
          both possibilities of the three sets. \\
        Cunningham uses only the first state set of the father node \cite{Cunningham1998}. \\
        From these two points four different versions of Fitch were formed:
        \begin{enumerate}
          \item Fitch 1: First state set of father node; intersection/union of siblings first.
          \item Fitch 2: First state set of father node; intersection/union of siblings together with father node.
          \item Fitch 3: Both state sets of father node; intersection/union of siblings first.
          \item Fitch 4: Both state sets of father node; intersection/union of siblings together with father node sets.
        \end{enumerate}
        \begin{figure}
          \centering
          \includegraphics[width=0.8\textwidth]{Figures/simulation_fitch_evaluation.png}
          \caption{Test of Fitch Versions.}
          \label{fig:Fitch versions}
        \end{figure}
        These four versions were tested in the simulation with 100 trees and 10000 leaf nodes and a
          distribution of 60~\% FL to 40~\% P. Figure \ref{fig:Fitch versions} shows this over all unknown 
          node percentage. \\
        % At 90~\% unknown nodes and 90~\% of multifurcation of the internal nodes, version 1 was 89.26~\%, 
        %   version 2 was 89.26~\%, version 3 was 89.35~\%, and version 4 was 89.31~\% correct. Therefore, 
        %   only version 3 was used for all further simulations.
        At 95~\% unknown nodes and 95~\% of multifurcation of the internal nodes, version 1 was 88.37~\%, 
          version 2 was 88.37~\%, version 3 was 88.4~\%, and version 4 was 88.39~\% correct. Therefore, 
          only version 3 was used for all further simulations.

      % Fitch versions:
      % | 88.37 % | 88.37 % | 88.4 % |88.39 % |

@@

    We examined the ridge of multifurcation of the tree. A complete phylogenetic tree would be 
      binary, which means the number of leaf nodes is closely to the number of internal nodes. But 
      since we only work with a synthesis tree, this tree is multifurcated: we have 241 974 internal 
      nodes and 2 293 463 leaf nodes. \\
    For a first overview we collected for every node its number of children (degree $-1$), and plotted
      this in two histograms, see figure \ref{fig:childrenOfNodes}. \\
    The multifurcation affects only the internal nodes. We collected the number of children $-2$ of 
      every node (a node with two children is binary). That means it discribes the number of nodes which we have lost from the real (binary) 
    phylogenetic tree. \\
    \begin{figure}[h!]
      \centering
      \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.9\textwidth]{Figures/multifurc.pdf}
        \caption{Histogram with automatic binwidth. \\ ~ \\ ~}
      \end{subfigure}
      ~~~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
        %(or a blank line to force the subfigure onto a new line)
      \begin{subfigure}[b]{0.42\textwidth}
        \includegraphics[trim = 0mm 0mm 30mm 0mm, clip, width=0.9\textwidth]{Figures/multifurc_small.pdf}
        \caption{Histogram with $binwidth = 1$. \\ light blue: binary; \\ dark blue: multifunction}
      \end{subfigure}
      \caption{Histograms about the multifurcation of the internal nodes of the synthesis tree.}
      \label{fig:childrenOfNodes}
    \end{figure}
    As \colorbox{red}{you} can see, we are very far from a binary tree. \\
    Some subtrees have been examined, these have the following percentages of missing information: See 
      table \ref{table:percentage loss information subtrees}.
    \begin{table}[h!]
      \begin{center}
        \begin{tabular}{ |l||r|r| }
          \hline
          Subtree of & unknown states & multifurcation \\ 
          \hline \hline
          Eukaryota       & 97.34~\%  & 98.16~\% \\
          \hline \hline
          Metazoa         & 96.44~\%  & 87.93~\% \\ \hline
          Fungi           & 98.87~\%  & 96.97~\% \\ \hline
          Chloroplastida  & 99.14~\%  & 89.46~\% \\
          \hline \hline            
          % Arthropoda      & 97.49~\%  & 89.95~\% \\ \hline
          Apicomplexa     & 86.26~\%  & 87.16~\% \\ \hline
          Nematoda        & 89.01~\%  & 88.59~\% \\ \hline
          Chordata        & 88.59~\%  & \cellcolor{green!50}66.49~\% \\ \hline
          Platyhelminthes & \cellcolor{green!50}68.73~\%  & 80.34~\% \\
          \hline \hline            
          Insecta         & 97.11~\%  & 90.78~\% \\
          \hline  
        \end{tabular}
      \end{center}
      \caption{Examination of subtrees regarding missing information. \\ 
        % Taxa of subtrees: Domain, Kingdom, Phylum, Class \\ 
        Unknown states: missing state information of leaf nodes \\ 
        Multifurcation: missing internal nodes}
      \label{table:percentage loss information subtrees} 
    \end{table}

@@

      % * Residuals: Fehler - wieviele Werte sind nicht gut modelliert. (umso kleiner umso besser - grün) \\
      The Residuals give us the error of the model. If the value is small, our data will be well modeled. \\
      Within the complexity classes it can be seen that the mean height gives the best additional factor. \\

@@

      Next to the problem of the multifurcation of the tree is the less of data we have for the species.
        For the ancestral state reconstruction, we need information in the leaf nodes. \\
      The eukaryotic synthesis tree has 293,463 leaf nodes. The GloBI database has 5 346 414 interactions 
        (at this timepoint). Out of this data we got 51,337 distinct free-living species and 47 332 
        distinct parasite species $\rightarrow$ unknown nodes 2,194,794 ($\approx 95.7~\%$). \\
      We found also 57,352 (not distinct!) source species and 809,993 (not distinct!) target
        species without OTT ids. Since we currently use only OTT ids, we could not use this information. \\
      % \todo{With this only $\approx 4.3~\%$ information in our leaf nodes are ...} \\

      We also compared different models in terms of their BICs (Table: \ref{table:BIC unknown information}). 
        The Residuals are not very meaningful here, since all models have different complexities. % /dimensions
        For the sake of completeness, the associated deviance tables are located with the residuals in 
        the appendix \ref{sec:Residuals unknown information}.

      \begin{table}[h!]
        \begin{center}
          \begin{tabular}{ |l|r|r|r|r| }
            \hline
            Model / Taxa & Kingdom & Phylum & Class & Order \\
            \hline \hline
            multifurc $\sim$ taxa & \cellcolor{green!15}545799 & \cellcolor{green!35}500004 & \cellcolor{green!45}485121 & XXXXX \\
            \hline
            multifurc $\sim$ taxa + depth & \cellcolor{green!15}544862 & \cellcolor{green!40}493808 & \cellcolor{green!45}481869 & \cellcolor{green!50}478851 \\
            \hline
            multifurc $\sim$ taxa * depth & \cellcolor{green!15}544179 & \cellcolor{green!45}489845 & \cellcolor{green!45}481494 & \cellcolor{green!50}478188 \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{BICs of unknown information}
        \label{table:BIC unknown information} 
      \end{table}
      It also follows from this table that the most complex model is the best. In general, the BIC 
        values are smaller than those of the multifurcation models. The modeling here is thus better. \\
      Again, the calculation of finer models (eg family) was too expensive. \\

      These missing data modeling results can be used to better simulate the data.

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Missing leaf state modelling - Resiudal tables}\label{sec:Residuals unknown information}

    \begin{table}[h!]
      \begin{center}
        \begin{tabular}{ |l|r|r|r| }
          \hline
          Model / Taxa & Kingdom & Phylum & Class \\
          \hline \hline
          multifurc $\sim$ taxa & 545740 & \cellcolor{green!10}499227 & \cellcolor{green!30}482265 \\
          \hline
          multifurc $\sim$ taxa + depth & 544789 & \cellcolor{green!10}493017 & \cellcolor{green!50}478998 \\
          \hline
          multifurc $\sim$ taxa * depth & 544062 & \cellcolor{green!30}488366 & \cellcolor{green!50}476382 \\
          \hline
        \end{tabular}
      \end{center}
      \caption{Residuals of unknown information}
    \end{table}
